>

### 2019-11-21 paper/spyrison-cook.Rmd
typos, edits, local function recency.

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional  Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, to provide a manual tour, that allows the coefficient for a single variable to be controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed using animation, with functions from either the \pkg{plotly} and \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. It is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour for studying the sensitivity of structure, in a projection, to specific variable contributions, is illustrated using particle physics data.

@@

```{r paperFunctions, include=FALSE}
## FUNCTION FOR STATIC OUTPPUT, 
# going to facets loses control of output size, simpler to just live with this.
array2static <- function(.m_tour, .data, .m_var, .cat, .mag =2.2)
{
  slides       <- array2df(array = .m_tour, data = .data)
  basis_slides <- slides$basis_slides
  data_slides  <- slides$data_slides
  n_slides     <- max(basis_slides$slide)
  if (n_slides != 15) 
    stop(paste0("stop: n_slides !=15 !!! Check the angle step size. n_slides = ", n_slides))

@@

  grid_b <- grid_t <-
    data.frame(slide = 1:n_slides, 
               x = .mag*rep(1:5, 3), y = .mag*rep(3:1, each = 5))
  grid_t$y <- grid_t$y + max(grid_t$y)
  # OUTER JOIN
  basis_grid <- merge(x = basis_slides, y = grid_t, by = "slide", all = TRUE)
  # CROSS JOIN
  circ_grid  <- merge(x = circ, y = grid_t, by = NULL)
  # OUTER JOIN
  data_grid  <- merge(x = data_slides, y = grid_b, by = "slide", all = TRUE)

  # BASIS AXES
  gg1 <-
    ggplot(data = basis_grid) +
    # AXES LINE SEGMETNS
    geom_segment(aes(x = V1 + x, y = V2 + y, xend = x, yend = y),
                 color = col_v, size = siz_v) +
    # AXES TEXT LABELS
    geom_text(aes(x = V1 + x, y = V2 + y, label = lab_abbr),
              color = col_v, vjust = "outward", hjust = "outward") +
    # AXES FRAME NUM
    geom_text(aes(x = x - .7, y = y + 1.1, 
                  label = paste0("frame: ",slide)), color = "grey50") +
    # AXES CIRCLE PATH
    suppressWarnings( # Suppress for "Removed 1 rows containing missing values."
      geom_path(data = circ_grid, color = "grey80",
                mapping = aes(x = x+c_x, y = y+c_y))
    )

  # PROJECTION
  gg2 <- gg1 +
    # PROJ DATA POINTS
    geom_point(data = data_grid, size = .7,
               mapping = aes(x = V1 + x, y = V2 + y, color = cat),
               shape = as.integer(cat) + 15) +
    # PROJ DATA FRAME NUM
    geom_text(data = data_grid, color = "grey50",
              mapping = aes(x = x - .7, y = y + 1.1, 
                            label = paste0("frame: ",slide))) +
    # OPTIONS
    theme_void() +
    scale_color_brewer(palette = "Dark2") + 
    theme(legend.position="none")
          #,panel.border = element_rect(colour = "black", fill = NA))

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions, such as principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach is to use a tour [@asimov_grand_1985] of dynamic linear projections, to look at many different low-dimensional projections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package \CRANpkg{tourr} [@wickham_tourr_2011] provides a platform for generating tours. It has many types of tours available, and many types of displays possible. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

@@

<!-- interesting features and guided tours -->
One way to identify "interesting" features is with the use of a guided tour [@cook_grand_1995]. Guided tours select a very specific path, which approaches a projection that optimizes an objective-function. The optimization is conducted in a manner similar to simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified an analyst can then use the "finer brush" of the manual tour, by controlling the contributions of individual variables to explore the sensitivity they have on the structure visible in the projection.

@@

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $\mathbb{R}^p \Rightarrow \mathbb{R}^2$. This is called this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".
<!--- $|\textbf{B}_k|$, the norm of the manip var, $k$, the length of this basis vector between 0 and 1 inclusive.
-->
- $\textbf{e}$, a 1D basis vector of length $p$, with 1 in the $k$-th position and 0 elsewhere.
- $\textbf{M}$ is a $p \times 3$ matrix, defining the 3D subspace where data rotation occurs and is called the manip(ulation) space.
- $\textbf{R}$, the 3D rotation matrix, for performing unconstrained 3D rotations within the manip space, $\textbf{M}$.
- $\theta$, the angle of in-projection rotation, for example, on the reference axes; $c_\theta, s_\theta$ are the cosine and sine.
- $\phi$, the angle of out-of-projection rotation, into the manip space; $c_\phi, s_\phi$ are the cosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for out-of-projection rotation orthogonal to $\textbf{e}$.
- $\textbf{Y}$, the resulting projection of the data through the manip space, $\textbf{M}$, and rotation matrix, $\textbf{R}$.
<!--
XXX Clean up the notation, not all of this is needed or useful. 
- The shadow of $e$ in the projection may need some notation. 
- R is really an unconstrained rotation? (constrained rotation mentioned twice, what is the difference?
-->

<!-- operate on bases -->
The algorithm operates entirely on projection bases and incorporates the data only when making the projected data plots.

@@

<!-- describe data. -->
The flea data (@lubischew_use_1962), available in the \pkg{tourr} package, is used to illustrate the algorithm. The data contains 74 observations on 6 variables, which are physical measurements made on beetles. Each observation belongs to one of three species.

<!-- Projection basis -->
An initial 2D projection basis must be provided. A suggested way to start is to identify an interesting projection using a projection pursuit guided tour. <!--In a guided tour, the projection sequence is selected by optimizing an index function via hill-climbing on the projection space.--> Here the holes index is used to find a 2D projection of the flea data, which shows three separated groups. Figure \ref{fig:step0} shows the initial projection of the data. The left panel displays the projection basis ($\textbf{B}$) and can be used as a visual guide of the magnitude and direction that each variable contributes to the projection. The right panel shows the projected data, $\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}$. The color and shape of points are mapped to the flea species. This plot is made using the \code{view\_basis()} function in \pkg{spinifex}, which generates a \CRANpkg{ggplot2} [@wickham_ggplot2:_2016] object.

@@

# step0, output
# Adjust centering to make axes and data side by side
view_basis(f_bas, labels = colnames(f_dat)) +
  geom_point(data = f_proj,
             mapping = aes(x = X1 + .75, y = X2 - .5, color = f_cat),
             pch = as.integer(f_cat) + 15) +
  theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

<!-- select a manip var-->
In Figure \ref{fig:step0} the contribution of the variables tars1 and aede2 mostly contrast the contribution of the other four variables. These two variables combined contribute in the direction of the projection where the purple cluster is separated from the other two clusters. The variable aede2 is selected as the manip var, the variable to be controlled in the tour. The question that will be explored, is, how important this variable is to the separation of the clusters.

@@

<!-- Zero Vect, manip sp -->
Initialize the coordinate basis vector as a zero vector $\textbf{e}$ of length $p$, and set the $k$-th element to 1. In the example data, aede2 is the fifth variable in the data, so $k=5$, set $e_5=1$. Use a Gram-Schmidt process to orthonormalize the coordinate basis vector on the original 2D projection to describe a 3D manip space, $\textbf{M}$.

@@

<!-- What the manip space provides -->
The manip space provides a 3D projection from $p$-dimensional space, where the coefficient of the manip var can range completely between [0, 1]. 3D rotation can be used to rotate the manip variable completely into or completely out of a 2D projection. Figure \ref{fig:step2} illustrates this 3D manip space with the manip var highlighted. This representation is produced by calling the \code{view\_manip\_space()} function. This diagram is purely used to help explain the algorithm.

```{r step2, echo=F, out.width='70%',                                           fig.cap = "Illustration of a 3D manip space, constructed to change the contribution of the variable aede2 in the example data. The red circle indicates a unit sphere. The 2D projection is represented by the blue circle, with the projection coefficients represented by grey lines and text. The manip var axis, in red, has length 1 touching the sphere, extends the projection to a third dimension. The shadow of this axis (blue) is its contribution in the original 2D projection. Illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, labels = colnames(f_dat)) +
    theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

<!-- illustration of axis manip -->
The basis vector corresponding to the manip var (red line in Figure \ref{fig:step2}), can be operated like a lever. This is the process of the manual control, that rotates the manip variable into and out of the 2D projection (Figure \ref{fig:step3}). As the variable contribution is controlled, the manip space rotates, and the projection onto the horizontal projection plane correspondingly changes. This is a manual tour. Generating a sequence of values for the rotation angles produces a path for the rotation of the manip space.

@@

<!-- Sequence of $\phi_i$ -->
1. Set initial value of $\phi_1$ and $\theta$: $\phi_1 = \cos^{-1}{\sqrt{B_{k1}^2+B_{k2}^2}}$, $\theta = \tan^{-1}\frac{B_{k2}}{B_{k1}}$. $\phi_1$ is the angle between $\textbf{e}$ and its shadow in $\textbf{B}$.
2. Set an angle increment ($\Delta_\phi$) that sets the step size for the animation, to rotate the manip var into and out of the projection. (Note: Using angle increment, rather than a number of steps, to control the movement, is consistent with the tour algorithm as implemented in the \pkg{tourr}).
3. Step towards $0$, where the manip var is completely in the projection.
4. Step towards $\pi/2$, where the manip variable has no contribution.
5. Step back to $\phi_1$.

In each of the steps 3-5, a small step may be added to ensure that the endpoints of $\phi$ ($0$, $\pi/2$) are reached.

@@

<!-- plot XY in seq for animation -->
\noindent where $\textbf{R}^{(i)}_{[3,3]}$ is the incremental rotation matrix, using $\phi_i$. To make the data plot, use the first two columns of \textbf{Y}. Show the projected data for each frame in sequence to form an animation. <!--The remaining manip space variable can be linked to point aesthetic (such as size or color) to produce depth cues used in conjunction with the $XY$ scatterplot.-->

<!-- Animation on flea data -->
Figure \ref{fig:step4} illustrates a manual tour sequence having 15 steps. The projection axes are displayed on the top half, which corresponds to the projected data in the bottom half. When aede2 is removed from the projection, the purple cluster overlaps with the green cluster. This suggests that aede2 is important for distinguishing this species.

@@

```{r step4, echo=F, warning=F, fig.height=7.0, fig.width=5.83, out.height='7in', out.width='5.83in',                                             fig.cap = "Radial manual tour manipulating aede2 of standardized flea data. The axis for aede2 increases in contribution to the projection, from its initial value to 1, decreasing to 0 and then returning to the initial value. This affects the separatiion between the purple and green clusters. This shows that aede2 is important for distinguishing the purple species, because the separation disappears when aede2 is not contributing to the projection. An animation can be viewed at https://nspyrison.netlify.com/thesis/flea\\_manualtour\\_mvar5/."}
#play_manual_tour(f_dat, f_bas, f_mvar, col = f_cat, angle = f_angle)
array2static(.m_tour = f_mtour, .data = f_dat, .m_var = f_mvar, .cat = f_cat)
```

@@

funcs <- rbind(
  c("construction", "create_manip_space", "forms the 3D space of rotation"),
  c("construction", "rotate_manip_space", "performs 3D rotation"),
  c("construction", "manual_tour", "generates sequence of 2D frames"),
  c("", "", ""),
  c("render", "render_", "constructs the ggplot object to feed to animation"),
  c("render", "render_plotly", "converts the ggplot object to a plotly animation"),
  c("render", "render_gganimate", "converts the ggplot object to a gganimate  animation"),
  c("render", "array2df", "turns a tour path array into long form, for plotting"),
  c("", "", ""),
  c("animation", "play_tour_path", "animates given tour path"),
  c("animation", "play_manual_tour", "animates the manual tour algorithm"),
  c("", "", ""),
  c("specialty", "view_basis", "displays the reference frame of a given basis"),
  c("specialty", "view_manip_space", "illustrative display of any manip space")
)
colnames(funcs) <- c("Type", "Function", "Description")

@@

<!-- setup -->
```{r step0CODEEXAMPLE, eval=F, echo=T}
library(spinifex)
f_data  <- tourr::rescale(flea[, 1:6])                    ## standardize data
f_path  <- save_history(f_data, guided_tour(holes()))     ## guided tour using holes index
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2) ## optimal projection
f_cat   <- factor(flea$species)                           ## categorical var
f_mvar  <- 5                                              ## set manip var
angle_speed <- .26
play_manual_tour(data = f_data,                           ## Generates animation
                 basis = f_basis,                         ## using plotly
                 manip_var = f_mvar, 
                 angle = angle_speed,
                 col = f_cat,
                 pch = f_cat)
```

The \code{play\_manual\_tour()} function is a composite function for user convenience that does these operations:

@@

The function `view_basis` can be used to show a projection, just the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

@@

<!-- step 2 disp code. -->
```{r step2CODEEXAMPLE, eval=F, echo=T}

view_manip_space(basis = f_basis, 
                 manip_var = f_mvar, 
                 labels = colnames(f_data))
```

<!--
## Rendering and sharing

The \pkg{tourr} package utilizes \pkg{base} graphics for the display of tours. \pkg{spinifex} allows tours to be rendered in \pkg{plotly} as an HTML5 object or \pkg{gganimate} as GIF or MP4 files. Sharing of animations is not trivial, especially static formats like print and PDF. Even with dynamic display capturing the correct resolution and aspect ratio can be challenging, while many formats quickly bloat file sizes limiting sharing options. Keep in mind hosting options and exporting functions offered in \pkg{plotly}, \pkg{gganimate} and \pkg{tourr}.

## Storage

Storing each data point for every frame of the animation is redundant. Just as operations are performed on the bases, so too should tour paths be stored as bases and a single instance of the data. Consider a radial manual tour, we can store the salient features in 3 bases, where $\phi$ is at its starting, minimum, and maximum values. The frames in between can be interpolated by supplying angular speed. With the use of the \code{tourr::save\_history()} function, the target bases can be saved. From there geodesic interpolation can be used to populate the intermittent frames. This type of interpolation should not be used on manual tours, which have already been initialized into a 3D manip space where direct linear interpolation is appropriate.
-->

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional  Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University 
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.

@@

```{r paperFunctions, include=FALSE}
## FUNCTION FOR STATIC OUTPPUT, 
# going to facets loses control of output size, simpler to just live with this.
array2static <- function(.m_tour, .data, .m_var, .cat, .mag = 2.2){
  ##' for @EXAMPLE 
  if (F){ #Example run:
    set.seed(20190425); f_dat <- tourr::rescale(flea[,1:6]); f_cat <- factor(flea$species);
    f_path <- c(.693, -.022, .082, -.119, .706, .023, -.070, .438, .405, .515, .103, .604); 
    f_bas <- matrix(f_path, ncol=2); f_mvar <- 5; f_angle <- .29;
    f_mtour <- manual_tour(f_bas, manip_var = f_mvar, angle = f_angle);
    .m_tour <- f_mtour; .data <- f_dat; .m_var <- f_mvar; .cat <- f_cat; .mag = 2.2;
  }

  slides       <- array2df(array = .m_tour, data = .data)
  basis_slides <- slides$basis_slides
  data_slides  <- slides$data_slides
  n_slides     <- max(basis_slides$slide)
  if (n_slides != 15) 
    stop(paste0("stop: n_slides !=15 !!! Check the angle step size. n_slides = ", n_slides))

@@

  grid_b <- grid_t <-
    data.frame(slide = 1:n_slides, 
               x = .mag*rep(1:5, 3), y = .mag*rep(3:1, each = 5))
  grid_t$y <- grid_t$y + max(grid_t$y)
  # OUTER JOIN
  basis_grid <- merge(x = basis_slides, y = grid_t, by = "slide", all = TRUE)
  ## grid_t$x becomes x.y . ie. colname.tbl
  # CROSS JOIN
  circ_grid  <- merge(x = circ, y = grid_t, by = NULL)
  # OUTER JOIN
  data_grid  <- merge(x = data_slides, y = grid_b, by = "slide", all = TRUE)

  # BASIS AXES
  gg1 <-
    ggplot(data = basis_grid) +
    # AXES LINE SEGMETNS
    geom_segment(aes(x = x.x + x.y, y = y.x + y.y, xend = x.y, yend = y.y),
                 color = col_v, size = siz_v) +
    # AXES TEXT LABELS
    geom_text(aes(x = x.x + x.y, y = y.x + y.y, label = lab),
              color = col_v, vjust = "outward", hjust = "outward") +
    # AXES FRAME NUM
    geom_text(aes(x = x.y - .7, y = y.y + 1.1, 
                  label = paste0("frame: ",slide)), color = "grey50") +
    # AXES CIRCLE PATH
    suppressWarnings( # Suppress for "Removed 1 rows containing missing values."
      geom_path(data = circ_grid, color = "grey80",
                mapping = aes(x = x+c_x, y = y+c_y))
    )

  # PROJECTION
  gg2 <- gg1 +
    # PROJ DATA POINTS
    geom_point(data = data_grid, size = .7,
               mapping = aes(x = x.x + x.y, y = y.x + y.y, color = cat),
               shape = as.integer(cat) + 15) +
    # PROJ DATA FRAME NUM
    geom_text(data = data_grid, color = "grey50",
              mapping = aes(x = x.y - .7, y = y.y + 1.1, 
                            label = paste0("frame: ",slide))) +
    # OPTIONS
    theme_void() +
    scale_color_brewer(palette = "Dark2") + 
    theme(legend.position="none")
          #,panel.border = element_rect(colour = "black", fill = NA))

@@

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions including principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach is to use a tour [@asimov_grand_1985] of dynamic linear projections to look at many different low-dimensional projections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package \CRANpkg{tourr} [@wickham_tourr_2011] provides a platform for generating tours. It can produce a variety of tours, each paried with a varitey of possible displays. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

@@

<!-- interesting features and guided tours -->
One way to identify "interesting" features is with the use of a guided tour [@cook_grand_1995]. Guided tours select a very specific path, which approaches a projection that optimizes an objective-function. The optimization is conducted in a manner similar to simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified, an analyst can then use the "finer brush" of the manual tour to controll the contributions of individual variables to explore the sensitivity they have on the structure visible in the projection.

@@

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $\mathbb{R}^p \Rightarrow \mathbb{R}^2$. This is called this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the variable to manipulate, called the "manip var".
<!--- $|\textbf{B}_k|$, the norm of the manip var, $k$, the length of this basis vector between 0 and 1 inclusive.
-->
- $\textbf{e}$, a 1D basis vector of length $p$, with 1 in the $k$-th position and 0 elsewhere.
- $\textbf{M}$ is a $p \times 3$ matrix, defining the 3D subspace where data rotation occurs and is called the manip(ulation) space.
- $\textbf{R}$, the 3D rotation matrix, for performing unconstrained 3D rotations within the manip space, $\textbf{M}$.
- $\theta$, the angle of in-projection rotation, for example, on the reference axes; $c_\theta, s_\theta$ are its cosine and sine.
- $\phi$, the angle of out-of-projection rotation, into the manip space; $c_\phi, s_\phi$ are itscosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for out-of-projection rotation orthogonal to $\textbf{e}$.
- $\textbf{Y}$, the resulting projection of the data through the manip space, $\textbf{M}$, and rotation matrix, $\textbf{R}$.
<!--
XXX Clean up the notation, not all of this is needed or useful. 
- The shadow of $e$ in the projection may need some notation. 
- R is really an unconstrained rotation? (constrained rotation mentioned twice, what is the difference?
-->

<!-- operate on bases -->
The algorithm operates entirely on projection bases and incorporates the data only when making the projected data plots, in light of efficiency.

@@

<!-- describe data. -->
The flea data (@lubischew_use_1962), available in the \pkg{tourr} package, is used to illustrate the algorithm. The data contains 74 observations on 6 variables, which are physical measurements made on flea beetles. Each observation belongs to one of three species.

<!-- Projection basis -->
An initial 2D projection basis must be provided. A suggested way to start is to identify an interesting projection using a projection pursuit guided tour. Here the holes index is used to find a 2D projection of the flea data, which shows three separated species groups. Figure \ref{fig:step0} shows the initial projection of the data. The left panel displays the projection basis ($\textbf{B}$) and can be used as a visual guide of the magnitude and direction that each variable contributes to the projection. The right panel shows the projected data, $\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}$. The color and shape of points are mapped to the flea species. This plot is made using the \code{view\_basis()} function in \pkg{spinifex}, which generates a \CRANpkg{ggplot2} [@wickham_ggplot2:_2016] object.

@@

# step0, output
# Adjust centering to make axes and data side by side
view_basis(f_bas, lab = colnames(f_dat)) +
  geom_point(data = f_proj,
             mapping = aes(x = X1 + .75, y = X2 - .5, color = f_cat),
             pch = as.integer(f_cat) + 15) +
  theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

<!-- select a manip var-->
In figure \ref{fig:step0} the contribution of the variables tars1 and aede2 mostly contrast the contribution of the other four variables. These two variables combined contribute in the direction of the projection where the purple cluster is separated from the other two clusters. The variable aede2 is selected as the manip var, the variable to be controlled in the tour. The question that will be explored is: how important is this variable to the separation of the clusters in this projection?

@@

<!-- Zero Vect, manip sp -->
Initialize the coordinate basis vector as a zero vector, $\textbf{e}$, of length $p$, and set the $k$-th element to 1. In the example data, aede2 is the fifth variable in the data, so $k=5$, set $e_5=1$. Use a Gram-Schmidt process to orthonormalize the coordinate basis vector on the original 2D projection to describe a 3D manip space, $\textbf{M}$.

@@

<!-- What the manip space provides -->
The manip space provides a 3D projection from $p$-dimensional space, where the coefficient of the manip var can range completely between [0, 1]. This 3D space serves as the medium that to rotate the projection basis relative to the selected manipulation variable. Figure \ref{fig:step2} illustrates this 3D manip space with the manip var highlighted. This representation is produced by calling the \code{view\_manip\_space()} function. This diagram is purely used to help explain the algorithm.

```{r step2, echo=F, out.width='70%',                                           fig.cap = "Illustration of a 3D manip space, this space is rotated effectively changing the contribution of the manip variable, aede2 in the example data. The blue circle and variable map lies on the projection plane. The red circle, orthogonal to the projection plane, illustrates the manipulation space and how the manip var can be controlled and how this effects the variable contribution back onto the projection plane. The other variables are omited from the manipulation dimension for simplicity. Picturing the other variables in that dimension reveals the intuition that rotating one variable performs a constrained rotation on the others. This is illustrated with the \\code{view\\_manip\\_space()} function."}
view_manip_space2(basis = f_bas, manip_var = f_mvar, labels = colnames(f_dat)) +
    theme(panel.border = element_rect(colour = "black", fill = NA))
```

@@

<!-- illustration of axis manip -->
The basis vector corresponding to the manip var (red line in Figure \ref{fig:step2}), can be operated like a lever anchored to the origin. This is the process of the manual control, that rotates the manip variable into and out of the 2D projection (Figure \ref{fig:step3}). As the variable contribution is controlled, the manip space rotates, and the projection onto the horizontal projection plane correspondingly changes. This is a manual tour. Generating a sequence of values for the rotation angles produces a path for the rotation of the manip space.

@@

<!-- Sequence of $\phi_i$ -->
1. Set initial value of $\phi_1$ and $\theta$: $\phi_1 = \cos^{-1}{\sqrt{B_{k1}^2+B_{k2}^2}}$, $\theta = \tan^{-1}\frac{B_{k2}}{B_{k1}}$. Where $\phi_1$ is the angle between $\textbf{e}$ and its shadow in $\textbf{B}$.
2. Set an angle increment ($\Delta_\phi$) that sets the step size for the animation, to rotate the manip var into and out of the projection. (Note: Using angle increment, rather than a number of steps, to control the movement, is consistent with the tour algorithm as implemented in the \pkg{tourr}).
3. Step towards $0$, where the manip var is completely in the projection plane.
4. Step towards $\pi/2$, where the manip variable has no contribution to the projection.
5. Step back to $\phi_1$.

In each of the steps 3-5, a small step may be added to ensure that the endpoints of $\phi$ ($0$, $\pi/2$, $\phi_1$) are reached.

@@

<!-- plot XY in seq for animation -->
\noindent where $\textbf{R}^{(i)}_{[3,3]}$ is the incremental rotation matrix, using $\phi_i$. To make the data plot, use the first two columns of \textbf{Y}. Show the projected data for each frame in sequence to form an animation.

<!-- Animation on flea data -->
Figure \ref{fig:step4} illustrates a manual tour sequence having 15 steps. The projection axes are displayed on the top half, which corresponds to the projected data in the bottom half. When aede2 is removed from the projection, the purple cluster overlaps with the green cluster. This suggests that aede2 is important for distinguishing between these species.

@@

```{r step4, echo=F, warning=F, fig.height=7.0, fig.width=5.83, out.height='7in', out.width='5.83in',                                             fig.cap = "Radial manual tour manipulating aede2 of standardized flea data. The axis for aede2 increases in contribution to the projection, from its initial value to 1, decreasing to 0 and then returning to the initial value. This effects the separation between the purple and green clusters. This shows that aede2 is important for distinguishing the purple species, because the separation disappears when aede2 is not contributing to the projection. An animation can be viewed at https://nspyrison.netlify.com/thesis/flea\\_manualtour\\_mvar5/."}
#play_manual_tour(f_dat, f_bas, f_mvar, col = f_cat, angle = f_angle)
array2static(.m_tour = f_mtour, .data = f_dat, .m_var = f_mvar, .cat = f_cat)
```

@@

funcs <- rbind(
  c("construction", "create_manip_space", "forms the 3D space of rotation"),
  c("construction", "rotate_manip_space", "performs 3D rotation"),
  c("construction", "manual_tour", "generates sequence of 2D frames"),
  c("", "", ""),
  c("render", "array2df", "turn the tour path array into long form, for plotting"),
  c("render", "render_", "render long form as a ggplot2 objection for animation"),
  c("render", "render_plotly", "render the animation as a plotly object (default)"),
  c("render", "render_gganimate", "render the animation as a gganimate object"),
  c("", "", ""),
  c("animation", "play_tour_path", "composite function animating the specified tour path"),
  c("animation", "play_manual_tour", "composite function animating the specified manual tour"),
  c("", "", ""),
  c("specialty", "view_basis", "display the reference frame of a given basis"),
  c("specialty", "view_manip_space", "illustrative display of any manip space")
)
colnames(funcs) <- c("Type", "Function", "Description")

@@

<!-- setup -->
```{r step0CODEEXAMPLE, eval=F, echo=T}
library(spinifex)
f_data  <- tourr::rescale(flea[, 1:6])                    ## standardized flea data
f_path  <- save_history(f_data, guided_tour(holes()))     ## guided tour path, holes index
f_basis <- matrix(f_path[,, max(dim(f_path)[3])], ncol=2) ## optimal projection
f_cat   <- factor(flea$species)                           ## categorical var
f_mvar  <- 5                                              ## set manip var number
angle_speed <- .26                                        ## anglular dist between slides
play_manual_tour(data = f_data,                           ## generates animation
                 basis = f_basis,                         ##   plotly obj (default)
                 manip_var = f_mvar, 
                 angle = angle_speed,
                 col = f_cat,
                 pch = f_cat)
```

\noindent The \code{play\_manual\_tour()} function is a composite function for user convenience that does these operations:

@@

The function `view_basis` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

@@

<!-- step 2 disp code. -->
```{r step2CODEEXAMPLE, eval=F, echo=T}
view_manip_space(basis = f_basis, 
                 manip_var = f_mvar, 
                 labels = colnames(f_data))
```
